An investigation into using scagnostics to find interesting projections of high-dimensional data

### 2018-10-05 scagnostics.Rmd
Small changes throughout.

* Information based index functions from @Reshef1518 are based on the mutual information $I$ when considering the distribution induced by the data points $D \subset \mathbb{R}^2$ on an $x-$by$y$ grid $G$. With $I^{\ast} (D, x, y) = \max I (D|_G)$ the characteristic matrix $M$ is defined as 
\begin{equation}
M(D)_{x,y} = \frac{I^{\ast} (D, x, y)}{\log \min{x,y}}.
\end{equation}
We can now define the following metrics:
    + Maximum Information Coefficient (MIC):
    \begin{equation}
    MIC(D) = \max_{xy<B(n)}\{M(D)_{x,y}\},
    \end{equation}
    where by default $B(n) = n^{0.6}$
    + Total Information Coefficient (TIC), see @JMLRv1715308:
    \begin{equation}
    TIC(D) = \sum_{xy<B(n)} M(D)_{x,y}
    \end{equation}
  While $MIC$ is normalised to fall between $[0,1]$, the upper bound of $TIC$ depends on the sample size via the bounding function $B(n)$. $TIC$ was designed to be more stable test against null hypothesis of independence, using the full information of the characteristic matrix rather than just the maximum, and as such is expected to be better suited as a ppi.

@@

```{r plannedtour, fig.width=6, fig.height=7, fig.cap="Short tour paths with scagnostics computed on projections. Most of the indices exhibit sharp jumps in scagnostic value."}
fullResMeltPlanned <- melt(fullResPlanned, id=c("t", "name", "size"))
ggplot(fullResMeltPlanned, aes(x=t, y=value)) +
  geom_line(aes(color=factor(size))) + 
  facet_wrap(variable~name, ncol=3, scales = "free_y", labeller = label_wrap_gen(multi_line=FALSE)) +
  guides(color=FALSE)
```

@@

@@

We next show in Fig \ref{fig:testpipe} the final view corresponding to the highest index found in the TIC scouting phase. While this is clearly not the optimal view of the pipe feature, we can already guess the type of dependence in the distribution, and we expect to be able to find the maximum when using this plane as a starting point of the geodesic optimisation. Indeed performing a standard guided tour with the TIC index and the identified starting plane efficiently identifies the optimal view shown in Fig ...
((Probably should remove second trace plot as it does not give much information, maybe show final view in same figure as final scouting view...))

```{r testpipe, out.width=".5\\textwidth", fig.cap="Final view discovered by the scouting phase."}
iLast <- length(pipeTourFull)
fProj <- pipeTourFull[[iLast]]
dProj <- as.tibble(pipeResc %*% fProj)
ggplot(dProj, aes(V1,V2)) + geom_point()
```

```{r refinepipe, results="hide"}
if(!file.exists("cache/refinepipe.rda")){
  pipeTour2 <- save_history(pipeResc, guided_tour(mineIndex("TIC")), start = fProj)
  pipeTourFull2 <- as.list(interpolate(pipeTour2))
  pipeTourRes2 <- getProj(pipe1000, pipeTourFull2, "Pipe", 1000)
  pipeTourResMelt2 <- melt(pipeTourRes2, id=c("t", "name", "size"))
  save(pipeTour2, pipeTourFull2, pipeTourRes2, pipeTourResMelt2, file = "cache/refinepipe.rda")
} else {
  load("cache/refinepipe.rda")
}
ggplot(pipeTourResMelt2, aes(x=t, y=value)) +
  geom_line() + 
  facet_wrap(~variable, ncol=1, scales = "free_y")

```

```{r testpipe2, out.width=".5\\textwidth"}
iLast <- length(pipeTourFull2)
fProj <- pipeTourFull2[[iLast]]
dProj <- as.tibble(pipeResc %*% fProj)
ggplot(dProj, aes(V1,V2)) + geom_point()
```

>>>>

* Information based index functions from @Reshef1518 are based on the mutual information $I$ when considering the distribution induced by the data points $D \subset \mathbb{R}^2$ on an $x-$by$y$ grid $G$. With $I^{\ast} (D, x, y) = \max I (D|_G)$ the characteristic matrix $M$ is defined as \footnote{We note that all descriptions are idealized, while in practice approximations and binning are used to obtain reasonable computing time, we refer to original references and package documentation for details.}
\begin{equation}
M(D)_{x,y} = \frac{I^{\ast} (D, x, y)}{\log \min{x,y}}.
\end{equation}
We can now define the following metrics:
    + Maximum Information Coefficient (MIC):
    \begin{equation}
    MIC(D) = \max_{xy<B(n)}\{M(D)_{x,y}\},
    \end{equation}
    where by default $B(n) = n^{0.6}$
    + Total Information Coefficient (TIC), see @JMLRv1715308:
    \begin{equation}
    TIC(D) = \sum_{xy<B(n)} M(D)_{x,y}
    \end{equation}
  While $MIC$ is normalised to fall between $[0,1]$, the upper bound of $TIC$ depends on the sample size via the bounding function $B(n)$. $TIC$ was designed to be more stable test against null hypothesis of independence, using the full information of the characteristic matrix rather than just the maximum, and as such is expected to be better suited as a ppi.

@@

```{r plannedtour, fig.width=6, fig.height=7, fig.cap="As Figure \ref{fig:plotV1V2toV3V4} but moving from the projection onto V1-V2 to the projection onto V5-V6, i.e. starting from random distribution view, and moving into the special view."}
fullResMeltPlanned <- melt(fullResPlanned, id=c("t", "name", "size"))
ggplot(fullResMeltPlanned, aes(x=t, y=value)) +
  geom_line(aes(color=factor(size))) + 
  facet_wrap(variable~name, ncol=3, scales = "free_y", labeller = label_wrap_gen(multi_line=FALSE)) +
  guides(color=FALSE)
```

## Distributions of index values

Compare typical index values over the two tour paths. Notation: path 1 / 2 labels tour from V1-V2 to V3-V4 / V5-V6. Typically find broader distributions for small samples because of the large fluctuations in this case. Some indices show preferred behaviour, i.e. sharp peaks around low values for path 1, and very broad distributions leading up to maximum values for path 2. For other indices we can see the strong dependence of the index on sample size, the most extreme example is the skinny index.

No clear interpretation without trace for comparison? See e.g. TIC for path2 pipe vs sine

```{r indexDist, fig.width=6, fig.height=7, fig.cap="Comparing distributions of index values found in the interpolated tour paths 1 and 2 as described in the text."}
fullResMelt1 <- fullResMelt %>% add_column(path="path1")
fullResMeltPlanned1 <- fullResMeltPlanned %>% add_column(path="path2")
allRes <- rbind(fullResMelt1, fullResMeltPlanned1)
ggplot(allRes, aes(x=value)) +
  geom_density(aes(color=path, linetype = factor(size))) +
  facet_wrap(variable~name, ncol=3, scales = "free_y", labeller = label_wrap_gen(multi_line=FALSE))
```

@@

There is some issue still with search\_better not actually giving monotonically increasing results, my guess is it is because of rigid assignment of x and y axis, i.e. calculating index on starting plane, compare to randomly selected new planes, if new index > starting index and with step size = inf (default when using save history) this means we move to the new plane, but in the interpolation framework, i.e. may be rotated wrt new plane found by search\_better, and since now the distribution is the same, but x/y values are not, we get somewhat different values, this may result in somewhat smaller values than anticipated in the new starting plane, which may result in small drops rather than monotonic increase.

@@

We next show in Fig \ref{fig:testpipe} (left) the final view corresponding to the highest index found in the TIC scouting phase. While this is clearly not the optimal view of the pipe feature, we can already identify the feature in the distribution, and we expect to be able to find the maximum when using this plane as a starting point of the geodesic optimisation. Indeed performing a standard guided tour with the TIC index and the identified starting plane efficiently identifies the optimal view shown in the right plot. In this case the optimisation quickly converges, following a smooth, monotonic increas of the TIC index. ((not shown because that is already full informtion basically))

```{r refinepipe, results="hide"}
if(!file.exists("cache/refinepipe.rda")){
  pipeTour2 <- save_history(pipeResc, guided_tour(mineIndex("TIC")), start = fProj)
  pipeTourFull2 <- as.list(interpolate(pipeTour2))
  pipeTourRes2 <- getProj(pipe1000, pipeTourFull2, "Pipe", 1000)
  pipeTourResMelt2 <- melt(pipeTourRes2, id=c("t", "name", "size"))
  save(pipeTour2, pipeTourFull2, pipeTourRes2, pipeTourResMelt2, file = "cache/refinepipe.rda")
} else {
  load("cache/refinepipe.rda")
}
```

```{r testpipe,  fig.height=4, fig.cap="Final view discovered by the scouting phase (left) and after second optimisation (right)."}
iLast <- length(pipeTourFull)
fProj <- pipeTourFull[[iLast]]
dProj <- as.tibble(pipeResc %*% fProj)
p1 <- ggplot(dProj, aes(V1,V2)) + geom_point()
iLast <- length(pipeTourFull2)
fProj <- pipeTourFull2[[iLast]]
dProj <- as.tibble(pipeResc %*% fProj)
p2 <- ggplot(dProj, aes(V1,V2)) + geom_point()
grid.arrange(p1, p2, ncol=2)
```
